From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-09 Pi+/paper/discussion.tex
Rest of styling
Small fixes to discussion

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible programming based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible languages.
Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence
suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels} and established in this work.

@@

In the remainder of this section, we discuss some broader related work.

@@

Higher-order term-rewriting systems and word problems have a long history of being formalised in proof assistants like
{homotopy.io}, Lean, Coq, and HoTT~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main
result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem and formalised its
correctness and termination. \citet{Hiver-coq} describes an explicit algorithm for producing normal forms. could provide
an alternative to our rewriting system. Other encodings of permutations as listed vectors, matrices, inductively
generated trees (Motzkin trees), Young diagrams, or string diagrams, proved either difficult formalise in type theory or
difficult to relate directly to the primitive type isomorphisms of $\PiLang$. The automatic
\citet{knuthSimpleWordProblems1970} completion produced too many equations making proving correctness and termination
intractable.

@@

There are other, different approaches to computing with univalence,
in~\cite*{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in
Cubical Type
Theory\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019},
univalence computes.

>>>>

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible programming based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible languages.
Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence
suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels} and established in this work. In the
remainder of this section, we discuss some broader related work.

@@

@@

Higher-order term-rewriting systems and word problems have a long history of being formalised in proof assistants like
{homotopy.io}, Agda, Coq and Lean~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main
result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem. \citet{Hiver-coq}
describes an explicit algorithm for producing normal forms. could provide an alternative to our rewriting system. Other
encodings of permutations as listed vectors, matrices, inductively generated trees (Motzkin trees), Young diagrams, or
string diagrams, proved either difficult formalise in type theory or difficult to relate directly to the primitive type
isomorphisms of $\PiLang$. The automatic \citet{knuthSimpleWordProblems1970} completion produced too many equations
making proving correctness and termination intractable.

@@

There are other, different approaches to computing with univalence,
in~\cite*{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in
Cubical Type
Theory\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.
